---
title: "Hotel Bookings Predictive Model"
---

## Overview

The data we are going to use is hotel bookings data. It came from Antonio, Almeida, and Nunes (2019). We will create a predictive model to predict which hotel stays included babies and/or children. We will use different hotel stay attributes to come up a prediction. Below is our data dictionary that should help us understand the data better.

+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| variable                       | class     | description                                                                                                                                                                                                                                                      |
+================================+===========+==================================================================================================================================================================================================================================================================+
| hotel                          | character | Hotel (H1 = Resort Hotel or H2 = City Hotel)                                                                                                                                                                                                                     |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| is_canceled                    | double    | Value indicating if the booking was canceled (1) or not (0)                                                                                                                                                                                                      |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| lead_time                      | double    | Number of days that elapsed between the entering date of the booking into the PMS and the arrival date                                                                                                                                                           |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| arrival_date_year              | double    | Year of arrival date                                                                                                                                                                                                                                             |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| arrival_date_month             | character | Month of arrival date                                                                                                                                                                                                                                            |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| arrival_date_week_number       | double    | Week number of year for arrival date                                                                                                                                                                                                                             |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| arrival_date_day_of_month      | double    | Day of arrival date                                                                                                                                                                                                                                              |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| stays_in_weekend_nights        | double    | Number of weekend nights (Saturday or Sunday) the guest stayed or booked to stay at the hotel                                                                                                                                                                    |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| stays_in_week_nights           | double    | Number of week nights (Monday to Friday) the guest stayed or booked to stay at the hotel                                                                                                                                                                         |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| adults                         | double    | Number of adults                                                                                                                                                                                                                                                 |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| children                       | double    | Number of children                                                                                                                                                                                                                                               |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| babies                         | double    | Number of babies                                                                                                                                                                                                                                                 |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| meal                           | character | Type of meal booked. Categories are presented in standard hospitality meal packages:\                                                                                                                                                                            |
|                                |           | Undefined/SC – no meal package;\                                                                                                                                                                                                                                 |
|                                |           | BB – Bed & Breakfast;\                                                                                                                                                                                                                                           |
|                                |           | HB – Half board (breakfast and one other meal – usually dinner);\                                                                                                                                                                                                |
|                                |           | FB – Full board (breakfast, lunch and dinner)                                                                                                                                                                                                                    |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| country                        | character | Country of origin. Categories are represented in the ISO 3155–3:2013 format                                                                                                                                                                                      |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| market_segment                 | character | Market segment designation. In categories, the term "TA" means "Travel Agents" and "TO" means "Tour Operators"                                                                                                                                                   |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| distribution_channel           | character | Booking distribution channel. The term "TA" means "Travel Agents" and "TO" means "Tour Operators"                                                                                                                                                                |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| is_repeated_guest              | double    | Value indicating if the booking name was from a repeated guest (1) or not (0)                                                                                                                                                                                    |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| previous_cancellations         | double    | Number of previous bookings that were cancelled by the customer prior to the current booking                                                                                                                                                                     |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| previous_bookings_not_canceled | double    | Number of previous bookings not cancelled by the customer prior to the current booking                                                                                                                                                                           |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| reserved_room_type             | character | Code of room type reserved. Code is presented instead of designation for anonymity reasons                                                                                                                                                                       |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| assigned_room_type             | character | Code for the type of room assigned to the booking. Sometimes the assigned room type differs from the reserved room type due to hotel operation reasons (e.g. overbooking) or by customer request. Code is presented instead of designation for anonymity reasons |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| booking_changes                | double    | Number of changes/amendments made to the booking from the moment the booking was entered on the PMS until the moment of check-in or cancellation                                                                                                                 |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| deposit_type                   | character | Indication on if the customer made a deposit to guarantee the booking. This variable can assume three categories:\                                                                                                                                               |
|                                |           | No Deposit – no deposit was made;\                                                                                                                                                                                                                               |
|                                |           | Non Refund – a deposit was made in the value of the total stay cost;\                                                                                                                                                                                            |
|                                |           | Refundable – a deposit was made with a value under the total cost of stay.                                                                                                                                                                                       |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| agent                          | character | ID of the travel agency that made the booking                                                                                                                                                                                                                    |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| company                        | character | ID of the company/entity that made the booking or responsible for paying the booking. ID is presented instead of designation for anonymity reasons                                                                                                               |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| days_in_waiting_list           | double    | Number of days the booking was in the waiting list before it was confirmed to the customer                                                                                                                                                                       |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| customer_type                  | character | Type of booking, assuming one of four categories:\                                                                                                                                                                                                               |
|                                |           | Contract - when the booking has an allotment or other type of contract associated to it;\                                                                                                                                                                        |
|                                |           | Group – when the booking is associated to a group;\                                                                                                                                                                                                              |
|                                |           | Transient – when the booking is not part of a group or contract, and is not associated to other transient booking;\                                                                                                                                              |
|                                |           | Transient-party – when the booking is transient, but is associated to at least other transient booking                                                                                                                                                           |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| adr                            | double    | Average Daily Rate as defined by dividing the sum of all lodging transactions by the total number of staying nights                                                                                                                                              |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| required_car_parking_spaces    | double    | Number of car parking spaces required by the customer                                                                                                                                                                                                            |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| total_of_special_requests      | double    | Number of special requests made by the customer (e.g. twin bed or high floor)                                                                                                                                                                                    |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| reservation_status             | character | Reservation last status, assuming one of three categories:\                                                                                                                                                                                                      |
|                                |           | Canceled – booking was canceled by the customer;\                                                                                                                                                                                                                |
|                                |           | Check-Out – customer has checked in but already departed;\                                                                                                                                                                                                       |
|                                |           | No-Show – customer did not check-in and did inform the hotel of the reason why                                                                                                                                                                                   |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| reservation_status_date        | double    | Date at which the last status was set. This variable can be used in conjunction with the ReservationStatus to understand when was the booking canceled or when did the customer checked-out of the hotel                                                         |
+--------------------------------+-----------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

## Library and Data Import

```{r warning=FALSE, message=FALSE}

library(tidymodels)
library(readr)

hotels = read_csv("https://tidymodels.org/start/case-study/hotels.csv") %>%
  mutate(across(where(is.character), as.factor))
```

## Outcome Variable Checking

Our goal is to build a motel that will predict which hotel stays included a children and/or babies and which did not. The column that will serve us our outcome variable is children.

```{r warning=FALSE, message=FALSE}

hotels %>% 
  count(children) %>% 
  mutate(
    prop = paste( round((n/sum(n)) * 100,2),"%",sep = "")
  )

```

We have here the proportion of hotel stay with and without children in it.

## Data Splitting and Resampling

We will use 70/30 split for our train and test data. Since we have an imbalance outcome variable class, we will use a stratified sample.

```{r warning=FALSE, message=FALSE}

set.seed(123)
splits = initial_split(hotels, strata = children)

hotel_train = training(splits)
hotel_test = testing(splits)


#Check Training set children proportion
hotel_train %>% 
  count(children) %>% 
  mutate(
    prop = n/sum(n)
  )


#Check Testing Set children proportion
hotel_test %>% 
  count(children) %>% 
  mutate(prop = n/sum(n))
```

We will use validation set resmpling method for our data splits.

```{r warning=FALSE, message=FALSE}

set.seed(234)

val_set = validation_split(hotel_train,
                           strata = children,
                           prop = 0.80)

val_set
```

## Model #1: Penalized Logistic Regression

Our outcome variable is a categorical column so its only reasonable for us to select Logistic Regression as our base model.

```{r warning=FALSE, message=FALSE}

lr_mod = logistic_reg(penalty = tune(), mixture = 1) %>% 
  set_engine("glmnet")
```

We set the `penalty` argument to `tune()` for now to serve as a placeholder temporarily. `mixture` was set to 1 to instruct our model to remove any possible nuisance variable keeping our model simple.

We can now proceed on creating our model recipe.

```{r warning=FALSE, message=FALSE}
holidays <- c("AllSouls", "AshWednesday", "ChristmasEve", "Easter", 
              "ChristmasDay", "GoodFriday", "NewYearsDay", "PalmSunday")

lr_recipe = recipe(children ~ ., data = hotel_train) %>% 
  step_date(arrival_date) %>% 
  step_holiday(arrival_date, holidays = holidays) %>% 
  step_rm(arrival_date) %>% 
  step_dummy(all_nominal_predictors()) %>% 
  step_zv(all_predictors()) %>% 
  step_normalize(all_predictors())

```

Creating Workflow

```{r warning=FALSE, message=FALSE}

lr_workflow = workflow() %>% 
  add_model(lr_mod) %>% 
  add_recipe(lr_recipe) 

```

Next, grid tuning object.

```{r warning=FALSE, message=FALSE}

lr_reg_grid <- tibble(penalty = 10^seq(-4, -1, length.out = 30))

lr_reg_grid %>% top_n(-5) # lowest penalty values

lr_reg_grid %>% top_n(5)  # highest penalty values


```

### Model Training and Tuning

```{r warning=FALSE, message=FALSE}

lr_res <- 
  lr_workflow %>% 
  tune_grid(val_set,
            grid = lr_reg_grid,
            control = control_grid(save_pred = TRUE),
            metrics = metric_set(roc_auc))


lr_plot <- 
  lr_res %>% 
  collect_metrics() %>% 
  ggplot(aes(x = penalty, y = mean)) + 
  geom_point() + 
  geom_line() + 
  ylab("Area under the ROC Curve") +
  scale_x_log10(labels = scales::label_number())

lr_plot 

```

This plots shows us that model performance is generally better at the smaller penalty values. This suggests that the majority of the predictors are important to the model. We also see a steep drop in the area under the ROC curve towards the highest penalty values. This happens because a large enough penalty will remove *all* predictors from the model, and not surprisingly predictive accuracy plummets with no predictors in the model (recall that an ROC AUC value of 0.50 means that the model does no better than chance at predicting the correct class).

```{r warning=FALSE, message=FALSE}

top_models <-
  lr_res %>% 
  show_best("roc_auc", n = 15) %>% 
  arrange(penalty) 
top_models

```

```{r warning=FALSE, message=FALSE}
lr_best <- 
  lr_res %>% 
  collect_metrics() %>% 
  arrange(penalty) %>% 
  slice(12)
lr_best
#> # A tibble: 1 × 7
#>   penalty .metric .estimator  mean     n std_err .config              
#>     <dbl> <chr>   <chr>      <dbl> <int>   <dbl> <chr>                
#> 1 0.00137 roc_auc binary     0.876     1      NA Preprocessor1_Model12
```

```{r warning=FALSE, message=FALSE}

lr_auc <- 
  lr_res %>% 
  collect_predictions(parameters = lr_best) %>% 
  roc_curve(children, .pred_children) %>% 
  mutate(model = "Logistic Regression")

autoplot(lr_auc)
```
